Paradigms in Physics 2.0

### 2017-06-29 paper.tex
comments from Mike

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee began with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students to for their
upper-division courses.
\end{abstract}

@@

\subsection{Math bits}
\davidsays{Discuss extra week and previous experience earlier.}
We chose to eliminate the \emph{Math Methods} course in favor of formal
\emph{Math Bits} in the \emph{Paradigm} courses.  Our previous \emph{Math Methods} course
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the \emph{Paradigm} courses, and allow advanced students to take our
graduate-level \emph{Math Methods} course.  In order to ensure that our
students to receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the \davidsays{``extra''
weeks having 3-week courses fitted into a 10-week quarter})
in the \emph{Paradigm} courses, we chose to integrate one week of \emph{Math Bits} into
each \emph{Paradigm} course.  The \emph{Math Bits} for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for
professors teaching \emph{Paradigm} courses, and at the same time ensures
that they do not succumb to the temptation to short-change the math content
to the detriment of the students.
\corinnesays{This paragraph is awkward.  I'm still thinking about some rewording.  This might be a good paragraph to discuss this afternoon.}

@@

\subsection{Computational lab}
Over the last six years, we have been developing a 1-credit
computational laboratory course that accompanies the \emph{Paradigm} courses.  We
chose to require this course, while removing a requirement for a
lower-division 3-credit course in computational physics.  This
lower-division course was challenging to teach, since it always had a
mix of lower- and upper-division students, with very different skill
levels and needs.  Transfer students now take computation alongside
the non-transfer students. The course is taught in a laboratory setting using pair
programming to help new programmers to learn.

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Physics of
Contemporary Challenges} and \emph{Techniques of Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The \emph{Challenges} course focuses on estimation,
dimensional reasoning, and interpretation of integrals, while
\emph{Theoretical Mechanics} teaches power series approximations and
exposes students to increased levels of mathematical sophistication.

@@

While contemporary challenges determine the narrative flow of the
course, physics concepts and physical reasoning skills are the main
substance. Students are introduced to thermodynamics, statistical
mechanics, electromagnetic radiation, quantum mechanics, and modern
experimental physics---in each case motivated by one or more
``contemporary challenges''. Together with new physics concepts,
students are given new reasoning tools, such as the equipartition
theorem, the quantum-classical correspondence principle, order of
magnitude calculations, simplifying assumptions, and numerical
integration. Mathematically rigorous derivations are only briefly
discussed in class, with additional details offered in online video
format. In summary, we emphasize the physics concepts and reasoning
skills that allow professional physicists to quickly/quantitatively
make an initial assessment of a complex problem.

@@

The \emph{Theoretical Mechanics} course is aimed at students who are taking
or have completed the last course in the introductory sequence. The
physics content of the course is advanced Newtonian mechanics,
introduction to Lagrangian and Hamiltonian techniques, and special
relativity. These topics at this stage for the students are convenient
for explicit discussion of sense-making strategies in physics because
(a) students are have recently studied problems that serve as limiting
cases for these more complex problems, (b) students at this level are
transitioning from solving problems primarily involving numbers to
problems with only symbolic parameters, and (c) sense-making as strategies 
can be discussed for approaching a problem, evaluating
answers, and refining intuitions about relativistic and other
unfamiliar situations. For example, we teach students how to use
spacetime diagrams to develop a story about a relativistic situation
and how to use hyperbola trigonometry on these diagrams to perform Lorentz
transformations that can be checked against the results of algebraic
Lorentz transformations.  The sense-making goals of the course are on
an equal footing with the physics goals, and sense-making is integrated
explicitly in all aspects of the course: exams, homework and during
in-class activities. Our aim is that students will develop
sense-making skills that will improve their learning in advanced
courses and will be valued by their upper-division course instructors,
research advisors, and future employers.

>>>>

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee began with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students for their
upper-division courses.
\end{abstract}

@@

\subsection{Math bits}
\davidsays{Discuss extra week and previous experience earlier.}
We chose to eliminate the \emph{Math Methods} course in favor of formal
\emph{Math Bits} in the \emph{Paradigm} courses.  Our previous \emph{Math Methods} course
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the \emph{Paradigm} courses, and allow advanced students to take our
graduate-level \emph{Math Methods} course.  In order to ensure that our
students receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the \davidsays{``extra''
weeks having 3-week courses fitted into a 10-week quarter})
in the \emph{Paradigm} courses, we chose to integrate one week of \emph{Math Bits} into
each \emph{Paradigm} course.  The \emph{Math Bits} for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for
professors teaching \emph{Paradigm} courses, and at the same time ensures
that they do not succumb to the temptation to short-change the math content
to the detriment of the students.
\corinnesays{This paragraph is awkward.  I'm still thinking about some rewording.  This might be a good paragraph to discuss this afternoon.}

@@

\subsection{Computational lab}
Over the last six years, we have been developing a 1-credit
computational laboratory course that accompanies the \emph{Paradigm} courses.  We
chose to require this course, while removing a requirement for a
lower-division 3-credit course in computational physics.  This
lower-division course was challenging to teach, since it always had a
mix of lower- and upper-division students, with very different skill
levels and needs.  Transfer students now take computation alongside
the non-transfer students. The course is taught in a laboratory setting using pair
programming to help new programmers to learn. \davidsays{CITE!}

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Physics of
Contemporary Challenges} and \emph{Techniques of Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The \emph{Challenges} course focuses on estimation,
dimensional reasoning, and interpretation of integrals, while
\emph{Theoretical Mechanics} teaches power series approximations and
exposes students to increased levels of mathematical sophistication. \davidsays{ADD CITATIONS SOMEWHERE HERE TO KELBY AND MACKENZIE}

@@

While contemporary challenges determine the narrative flow of the
course, physics concepts and physical reasoning skills are the main
substance. Students are introduced to thermodynamics, statistical
mechanics, electromagnetic radiation, quantum mechanics, and modern
experimental physics---in each case motivated by one or more
``contemporary challenges.'' Together with new physics concepts,
students are given new reasoning tools, such as the equipartition
theorem, the quantum-classical correspondence principle, order of
magnitude calculations, simplifying assumptions, and numerical
integration. Mathematically rigorous derivations are only briefly
discussed in class, with additional details offered in online video
format. In summary, we emphasize the physics concepts and reasoning
skills that allow professional physicists to quickly/quantitatively
make an initial assessment of a complex problem.

@@

The \emph{Theoretical Mechanics} course is aimed at students who are taking
or have completed the last course in the introductory sequence. The
physics content of the course is advanced Newtonian mechanics,
introduction to Lagrangian and Hamiltonian techniques, and special
relativity. These topics at this stage for the students are convenient
for explicit discussion of sense-making strategies in physics because
(a) students are have recently studied problems that serve as limiting
cases for these more complex problems, (b) students at this level are
transitioning from solving problems primarily involving numbers to
problems with only symbolic parameters, and (c) sense-making as strategies 
can be discussed for approaching a problem, evaluating
answers, and refining intuitions about relativistic and other
unfamiliar situations. For example, we teach students how to use
spacetime diagrams to develop a story about a relativistic situation
and how to use hyperbola trigonometry on these diagrams to perform Lorentz
transformations that can be checked against the results of algebraic
Lorentz transformations.  The sense-making goals of the course are on
an equal footing with the physics goals, and sense-making is integrated
explicitly in all aspects of the course: exams, homework and during
in-class activities. \davidsays{CITE HERE?} Our aim is that students will develop
sense-making skills that will improve their learning in advanced
courses and will be valued by their upper-division course instructors,
research advisors, and future employers.
